On the spatial aggregation of ecosystem condition indicators

### 2024-11-25 manuscript/manuscript.qmd
inc base size

---
title: On the spatial aggregation of ecosystem condition indicators
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    orchid: https://orcid.org/0000-0002-9623-9491
    affiliations: 
        - id: nina
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Pb 5685 Torgarden
          city: Trondheim
          postal-code: 7485
    attributes:
        corresponding: true
  - name: Hanno Sandvik
    email: hanno.sandvik@nina.no
    orchid: https://orcid.org/0000-0002-5889-1606
    affiliations:
      - id: nina
  - name: Bálint Czúcz
    email: balint.czucz@nina.no
    affiliations:
        - ref: nina
abstract: |
  Ecosystem condition assessments (ECA) and accounts use variables and indicators to describe key ecosystem characteristics reflecting the ecosystem condition and any potential deviations from a reference condition. These metrices are rutinely aggregation spatially to produce values representing the condition of a larger area, such as a country. However, we experience little awareness about the potential pitfals arising from aggregation bias of highly modified indicators. Here we outline some consequences regarding the order of the steps involed in normalising variables and aggregating them in space. We show that the choice of  aggregation pathways in non-trivial and has the potential to undermine the credibility and precition in ECAs, and to confuse the communication of their results. We introduce a common terminology for aggregation pathways, spesific to ecosystem condition indicators following SEEA EA standards, and make some recomendations about which pathway to use in different settings, and how to report these choices.

>>>>

---
title: On the spatial aggregation of ecosystem condition indicators
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    orchid: https://orcid.org/0000-0002-9623-9491
    affiliations: 
        - id: nina
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Pb 5685 Torgarden
          city: Trondheim
          postal-code: 7485
    attributes:
        corresponding: true
  - name: Hanno Sandvik
    email: hanno.sandvik@nina.no
    orchid: https://orcid.org/0000-0002-5889-1606
    affiliations:
      - ref: nina
  - name: Bálint Czúcz
    email: balint.czucz@nina.no
    affiliations:
      - ref: nina
abstract: |
  Ecosystem condition assessments (ECA) and accounts use variables and indicators to describe key ecosystem characteristics reflecting the ecosystem condition and any potential deviations from a reference condition. These metrices are rutinely aggregation spatially to produce values representing the condition of a larger area, such as a country. However, we experience little awareness about the potential pitfals arising from aggregation bias of highly modified indicators. Here we outline some consequences regarding the order of the steps involed in normalising variables and aggregating them in space. We show that the choice of  aggregation pathways in non-trivial and has the potential to undermine the credibility and precition in ECAs, and to confuse the communication of their results. We introduce a common terminology for aggregation pathways, spesific to ecosystem condition indicators following SEEA EA standards, and make some recomendations about which pathway to use in different settings, and how to report these choices.
